# (PART\*) Statistical Models {-}

# Ordinary Least Squares

- we want to solve something of the form
$$ \boldsymbol{Y} = \boldsymbol{X}\boldsymbol{\beta} + \boldsymbol{\epsilon} $$

## Solutions

- OLS is typically learned as

$$ (\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\boldsymbol{Y} $$
- but in practice, this is computationally inefficient
- floating point algebra, matrix inversion, etc etc
- instead, we use QR decomposition

### QR Decomposition

$$ \boldsymbol{X} = \boldsymbol{QR} $$
- since R is an upper triangular matrix, we can set

$$ \boldsymbol{Y} = \boldsymbol{QR}\boldsymbol{\beta} $$
- meaning that:

$$ \boldsymbol{Q}^{-1}\boldsymbol{Y} = \boldsymbol{R}\boldsymbol{\beta} $$
- by definition of the QR decomposition:
$$ \boldsymbol{Q}^{-1}\boldsymbol{Y} = \boldsymbol{Q}'\boldsymbol{Y}$$
- we can therefore exploit the triangular structur of $\boldsymbol{R}$ to backsolve the equation:
$$ \boldsymbol{Q}'\boldsymbol{Y} = \boldsymbol{R}\boldsymbol{\beta} $$

### LU Decomposition

$$ \left(\boldsymbol{X}'\boldsymbol{X}\right)\boldsymbol{b} = \boldsymbol{X}'\boldsymbol{y} $$



## Implementation


```{r}

ols <- function(data, formula, method="QR") {
  
  mf <- model.frame(formula, data, na.action=na.exclude)
  Y <- model.response(mf, type="double")
  X <- model.matrix(formula, mf)
  
  if (method == "matrix") b <- solve(crossprod(X)) %*% crossprod(X, Y)
  if (method == "QR") {
    QR <- qr(X)
    Q <- qr.Q(QR)
    R <- qr.R(QR)
    b <- backsolve(R, crossprod(Q,y))
    rownames(b) <- colnames(X)
  }
  
  return(b)
}

```

Testing the function:

```{r}

set.seed(42)

N <- 1000
X1 <- rnorm(N)
X2 <- rnorm(N)
y = 1.5 + 2*X1 + 3*X2 + rnorm(N)

df <- data.frame(
  y, X1, X2
)

f <- y ~ X1 + X2

coef(lm(f, df))
ols(df, f)
ols(df, f, "matrix")

```


